---
title: "SMT Data Challenge"
date: "2026-06-23"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)

source('SMT_Data_Starter.R')
```


```{r setup, message=FALSE, warning=FALSE}
# set coordinates for bases
SECOND_BASE <- c(x = 0, y = 90 * sqrt(2))   # (0, 127.28) ft
FIRST_BASE  <- c(x = 45 * sqrt(2), y = 45 * sqrt(2))
```

## Candidate plays from lineups

The following must be true: runner is on 1B, 2B empty, not a pickoff play. Runner on 3B is allowed but flagged (1st and 3rd steal dynamics differ). We also compute next-play context with continuity guards so we will not compare across half innings.

```{r candidates-lineups}
# lineups.csv is NOT always unique on (game_string, play_per_game): a small
# number of games, 43, have consecutive at-bats logged under the same
# play_per_game id, plus some exact duplicate rows. player_positions/ball_events
# only key on (game_string, play_per_game), so that pair must be unique here or
# every downstream join, especially later when using the runner-tracking join, becomes
# many-to-many and duplicates rows. Keep the latest at_bat per key,
# since it's the state immediately preceding the physically tracked play.

lineups_clean <- lineups %>%
  mutate(play_per_game = as.numeric(play_per_game),
         at_bat        = as.numeric(at_bat)) %>%
  arrange(game_string, play_per_game, at_bat) %>%
  group_by(game_string, play_per_game) %>%
  slice_max(at_bat, n = 1, with_ties = FALSE) %>%
  ungroup()

candidates <- lineups_clean %>%
  group_by(game_string) %>%
  mutate(
    next_on_1b       = lead(on_1b),
    next_on_2b       = lead(on_2b),
    next_on_3b       = lead(on_3b),
    next_at_bat      = lead(at_bat),
    next_half_inning = lead(half_inning),
    next_inning      = lead(inning),
    is_last_play_of_half = is.na(next_half_inning) |
                           next_half_inning != half_inning |
                           next_inning != inning
  ) %>%
  ungroup() %>%
  filter(!is.na(on_1b), is.na(on_2b), !is_pickoff) %>%
  mutate(runner_on_3b_flag = !is.na(on_3b))

nrow(candidates)
```

## Per-play event summaries from ball_events
Require a pitch, event code 1

and

Exclude: 
-ball hit into play, event code 4
-pickoff throws, event code 6

Catcher throw is kept as a feature (contested vs. uncontested), not a
filter because that would silently drop uncontested steals.

```{r events-per-play}
candidate_games <- unique(candidates$game_string)


events_raw <- ball_events %>%
  filter(game_string %in% candidate_games) %>%
  collect() %>%
  mutate(across(c(play_per_game, timestamp, player_id, ball_eventcode),
                as.numeric))

events_per_play <- events_raw %>%
  group_by(game_string, play_per_game) %>%
  summarize(
    event_codes        = str_c(ball_eventcode[order(timestamp)], collapse = ","),
    has_pitch          = any(ball_eventcode == 1),
    ball_hit_into_play = any(ball_eventcode == 4),
    pickoff_throw      = any(ball_eventcode == 6),
    pitch_timestamp    = suppressWarnings(min(timestamp[ball_eventcode == 1])),
    # catcher, player_id, acquiring the pitch, then throwing (event 3 then 8)
    catcher_acquire_ts = suppressWarnings(
                           min(timestamp[ball_eventcode %in% c(2, 7) &
                                         player_id == 2])),
    catcher_throw      = any(ball_eventcode %in% c(3, 8) & player_id == 2),
    catcher_throw_ts   = suppressWarnings(
                           min(timestamp[ball_eventcode %in% c(3, 8) &
                                         player_id == 2])),
    #middle infielder, 2B=4 or SS=6, receiving = someone covering the bag
    acquired_by_mif    = any(ball_eventcode == 2 & player_id %in% c(4, 6)),
    .groups = "drop"
  ) %>%
  mutate(across(c(pitch_timestamp, catcher_acquire_ts, catcher_throw_ts),
                ~ if_else(is.infinite(.x), NA_real_, .x)))

candidates <- candidates %>%
  left_join(events_per_play, by = c("game_string", "play_per_game")) %>%
  filter(has_pitch,
         !ball_hit_into_play | is.na(ball_hit_into_play),
         !pickoff_throw      | is.na(pickoff_throw))


nrow(candidates)
```

## Track runner movement (player_id 11 = runner on 1B)

Confirm a genuine attempt using progress toward 2B and an engineered variable called break timing. Break timing is how long after the pitch has the runner on 1B committed to a real sprint towards 2nd.
These summaries double as ML features.

```{r runner-movement}
runner_tracks <- player_positions %>%
  filter(game_string %in% candidate_games) %>%
  collect() %>%
  mutate(across(c(play_per_game, timestamp, player_id, field_x, field_y),
                as.numeric)) %>%
  filter(player_id == 11) %>%
  inner_join(candidates %>%
               select(game_string, play_per_game, pitch_timestamp,
                      catcher_acquire_ts),
             by = c("game_string", "play_per_game"))


runner_summary <- runner_tracks %>%
  filter(!is.na(pitch_timestamp)) %>%
  arrange(game_string, play_per_game, timestamp) %>%
  group_by(game_string, play_per_game) %>%
  mutate(
    dist_to_2b = sqrt((field_x - SECOND_BASE["x"])^2 +
                      (field_y - SECOND_BASE["y"])^2),
    dt         = (timestamp - lag(timestamp)) / 1000,          # seconds
    # speed of approach toward 2B (ft/s); positive = closing on the bag
    closing_speed = -(dist_to_2b - lag(dist_to_2b)) / dt
  ) %>%
  summarize(
    # lead distance: distance from 1B at pitch release
    lead_dist_at_pitch = sqrt(
      (first(field_x[timestamp >= pitch_timestamp]) - FIRST_BASE["x"])^2 +
      (first(field_y[timestamp >= pitch_timestamp]) - FIRST_BASE["y"])^2),
    # 1500 msec or 1.5 sec is an arbitrary time used to detect when a runner got a jump 1.5 secs before a pitch
    # relied on domain knowledge for it
    min_dist_to_2b  = min(dist_to_2b[timestamp >= pitch_timestamp - 1500],
                          na.rm = TRUE),
    # break time: first sustained sprint toward 2B (>= 12 ft/s = 8.2mph),
    # in seconds relative to pitch release (negative = broke before release)
    # 12 was chosen to catch the runners start of acceleration
    break_time = (suppressWarnings(
                    min(timestamp[closing_speed >= 12 &
                                  timestamp >= pitch_timestamp - 1500],
                        na.rm = TRUE)) - first(pitch_timestamp)) / 1000,
    max_speed_to_2b = max(closing_speed[timestamp >= pitch_timestamp - 1500],
                          na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ if_else(is.infinite(.x), NA_real_, .x)))

candidates <- candidates %>%
  left_join(runner_summary, by = c("game_string", "play_per_game"))
```

## Outcome labeling and attempt confirmation

* Excluding walks/HBP
* Inning-ending caught stealing plays get their own label instead
  of a cross-half-inning misread.
* Wild pitch/Pass ball separation: a runner who breaks only after the catcher already
  has (or misses) the ball advanced on a wild pitch / passed ball are not considered steals.

```{r steal-attempts}
all_steal_outcomes <- candidates %>%
  mutate(
    walk_like = !is_last_play_of_half &
                !is.na(next_on_1b) & next_on_1b == batter &
                !is.na(next_on_2b) & next_on_2b == on_1b,

    outcome = case_when(
      walk_like                                          ~ "Walk/HBP advance",
      !is_last_play_of_half & !is.na(next_on_2b) &
        next_on_2b == on_1b                              ~ "Safe",
      !is_last_play_of_half &
        (is.na(next_on_1b) | next_on_1b != on_1b) &
        (is.na(next_on_2b) | next_on_2b != on_1b) &
        (is.na(next_on_3b) | next_on_3b != on_1b) &
        next_at_bat == at_bat                            ~ "Out",
      is_last_play_of_half                               ~ "Out (inning-ending, verify)",
      TRUE                                               ~ "Ambiguous"
    ),

    # Genuine attempt: got close to 2B AND broke near pitch release
    confirmed_attempt = !is.na(min_dist_to_2b) & min_dist_to_2b < 30 &
                        !is.na(break_time)     & break_time < 1.0,

    # Break well AFTER catcher acquisition -> WP/PB advance, not a steal
    likely_wp_pb = !is.na(break_time) & !is.na(catcher_acquire_ts) &
                   !is.na(pitch_timestamp) &
                   (pitch_timestamp + break_time * 1000) > catcher_acquire_ts + 300,

    contested = catcher_throw & acquired_by_mif
  )

steal_attempts <- all_steal_outcomes %>%
  filter(outcome %in% c("Safe", "Out", "Out (inning-ending, verify)"),
         confirmed_attempt,
         !likely_wp_pb) %>%
  select(game_string, play_per_game, inning, half_inning, at_bat,
         runner = on_1b, batter, pitcher, catcher,
         outcome, contested, runner_on_3b_flag,
         lead_dist_at_pitch, break_time, min_dist_to_2b, max_speed_to_2b,
         catcher_throw, event_codes)

steal_attempts
```

## Diagnostics

```{r diagnostics}
#checking to make sure numbers look correct before finalizing steal attempts
cat("Candidate plays (base state + event filters):",
    nrow(candidates), "\n")
cat("Confirmed steal attempts of 2B:", nrow(steal_attempts), "\n\n")

steal_attempts %>% count(outcome, contested)

cat("\nBreak time (sec relative to pitch release):\n")
summary(steal_attempts$break_time)

cat("\nLead distance at pitch (ft):\n")
summary(steal_attempts$lead_dist_at_pitch)

#break_time should be roughly bimodal if WP/PB
# advances are sneaking through where real steals cluster near 0 and WP/PB well after.
ggplot(steal_attempts, aes(break_time, fill = outcome)) +
  geom_histogram(binwidth = 0.1, alpha = 0.7, position = "identity") +
  labs(title = "Runner break time relative to pitch release",
       x = "Break time (sec)", 
       y = "Plays")

```

## Spot-check ambiguous plays


```{r spot-check, eval=FALSE}
# Eyeball inning-ending CS labels and any borderline plays before training:
to_verify <- steal_attempts %>%
  filter(outcome == "Out (inning-ending, verify)")

#animate_play(to_verify$game_string[], to_verify$play_per_game[])
```
Went through each of the ambiguous edge case plays by hand and checked if they were steals or not.
Suspect plays:
index/play: 21/163, 28/414, 29/53, 43/189

## Animate a specific play 

```{r animate-row-102, eval=FALSE}
animate_play(steal_attempts$game_string[102], steal_attempts$play_per_game[102])
```

## Apply manual review corrections

Findings from watching `animate_play()` on each `to_verify` row above.
Keyed on (game_string, play_per_game) rather than row index, since
`to_verify`'s row order isn't stable across reruns.

```{r manual-corrections}
manual_review <- tribble(
  ~game_string,       ~play_per_game, ~manual_outcome,
  "y1_d145_ALC_ARN",   163,           "Out",  # confirmed caught stealing
  "y1_d163_SZJ_ANI",   414,           "exclude",  # 1st-and-3rd break; throw got lead runner at 3B, not a 1st->2nd steal
  "y1_d164.5_SZJ_ANI", 53,            "Out",      # confirmed caught stealing
  "y1_d212_QHX_ANI",   189,           "exclude"
)

steal_attempts <- steal_attempts %>%
  left_join(manual_review, by = c("game_string", "play_per_game")) %>%
  mutate(outcome = coalesce(manual_outcome, outcome)) %>%
  filter(outcome != "exclude") %>%
  select(-manual_outcome)

cat("Confirmed steal attempts of 2B after manual review:", nrow(steal_attempts), "\n")
```


```{r manual-corrections}
steal_attempts %>% count(outcome, contested)
```

```{r}
saveRDS(steal_attempts, "steal_attempts.rds") 
saveRDS(runner_tracks, "runner_tracks.rds")
```

