---
title: "Catcher Info: Pop Time & Throw Speed"
date: "2026-07-08"
output: html_document
---

This file builds `catcher_info` — pop time, exchange time, and throw speed for
the catcher's throw to second base on each steal-attempt play.

**Prerequisite:** run `all_steal_outcomes.Rmd` first. This file reuses objects
already in the environment from that workflow:

- `steal_attempts` — confirmed attempts with final Safe/Out labels (post manual review)
- `SECOND_BASE` — second base coordinates
- `ball_events`, `ball_positions` — Arrow datasets from `SMT_Data_Starter.R`

`steal_plays` and `steal_games` are derived below directly from `steal_attempts`,
not from the broader (unconfirmed/ambiguous) `all_steal_outcomes` table.

Event logic per play: pitch (code 1) → catcher receives (code 2, player_id 2) →
catcher throws (code 3, player_id 2) → middle infielder receives at the bag
(code 2, player_id 4 or 6). Pop time = catcher receipt → infielder receipt.

```{r key-event-times}
# Scope to confirmed attempts only (Safe/Out, post manual review) rather than
# all_steal_outcomes -- that table still carries Ambiguous/Walk-HBP/excluded
# rows, which would drag irrelevant plays into the throw-tracking join below
# and make the sanity-check counts read as "all candidates" instead of
# "confirmed steals."
steal_plays <- steal_attempts %>%
  distinct(game_string, play_per_game, outcome)
steal_games <- unique(steal_plays$game_string)

# Ball events for the steal plays only
steal_events <- ball_events %>%
  filter(game_string %in% steal_games) %>%
  collect() %>%
  mutate(across(c(play_per_game, timestamp, player_id, ball_eventcode),
                as.numeric)) %>%
  semi_join(steal_plays, by = c("game_string", "play_per_game")) %>%
  arrange(game_string, play_per_game, timestamp)

# Helper: earliest timestamp meeting a condition, NA if none
first_time <- function(ts, cond) {
  hits <- ts[cond & !is.na(cond)]
  if (length(hits) == 0) NA_real_ else min(hits)
}

# Key event timestamps per play
event_times <- steal_events %>%
  group_by(game_string, play_per_game) %>%
  summarize(
    pitch_time   = first_time(timestamp, ball_eventcode == 1),
    # catcher receives the pitch
    catch_time   = first_time(timestamp,
                              ball_eventcode == 2 & player_id == 2 &
                                timestamp >= pitch_time),
    # catcher releases the throw
    throw_time   = first_time(timestamp,
                              ball_eventcode == 3 & player_id == 2 &
                                timestamp >= catch_time),
    # 2B (4) or SS (6) receives the throw at the bag
    receive_time = first_time(timestamp,
                              ball_eventcode == 2 & player_id %in% c(4, 6) &
                                timestamp >= throw_time),
    .groups = "drop"
  )
```

```{r throw-tracking}
# Ball tracking during the throw window (release -> receipt)
throw_plays <- event_times %>%
  filter(!is.na(throw_time), !is.na(receive_time)) %>%
  # guard: one row per play so the tracking join can't double-count frames
  distinct(game_string, play_per_game, .keep_all = TRUE)

throw_metrics <- ball_positions %>%
  filter(game_string %in% unique(throw_plays$game_string)) %>%
  collect() %>%
  mutate(across(c(play_per_game, timestamp, ball_position_x,
                  ball_position_y, ball_position_z), as.numeric)) %>%
  inner_join(throw_plays, by = c("game_string", "play_per_game")) %>%
  filter(timestamp >= throw_time, timestamp <= receive_time) %>%
  arrange(game_string, play_per_game, timestamp) %>%
  group_by(game_string, play_per_game) %>%
  mutate(
    dt        = (timestamp - lag(timestamp)) / 1000,          # seconds
    step_dist = sqrt((ball_position_x - lag(ball_position_x))^2 +
                     (ball_position_y - lag(ball_position_y))^2 +
                     (ball_position_z - lag(ball_position_z))^2),
    speed_mph = if_else(dt > 0, (step_dist / dt) * 0.681818, NA_real_)
  ) %>%
  summarize(
    max_throw_speed_mph = max(speed_mph, na.rm = TRUE),
    p95_throw_speed_mph = quantile(speed_mph, 0.95, na.rm = TRUE),
    avg_throw_speed_mph = 0.681818 * sum(step_dist, na.rm = TRUE) /
                            ((max(timestamp) - min(timestamp)) / 1000),
    arrival_dist_to_2b  = sqrt((last(ball_position_x) - SECOND_BASE["x"])^2 +
                               (last(ball_position_y) - SECOND_BASE["y"])^2),
    n_frames            = n(),
    .groups = "drop"
  )
```

```{r assemble-catcher-info}
catcher_info <- steal_plays %>%
  left_join(event_times,   by = c("game_string", "play_per_game")) %>%
  left_join(throw_metrics, by = c("game_string", "play_per_game")) %>%
  mutate(
    pop_time      = (receive_time - catch_time) / 1000,   # catch -> tag receipt
    exchange_time = (throw_time   - catch_time) / 1000,   # catch -> release
    flight_time   = (receive_time - throw_time) / 1000,   # release -> receipt
    throw_to_2b   = !is.na(arrival_dist_to_2b) & arrival_dist_to_2b <= 15
  ) %>%
  select(game_string, play_per_game, outcome, pop_time, exchange_time,
         flight_time, max_throw_speed_mph, p95_throw_speed_mph,
         avg_throw_speed_mph, throw_to_2b, arrival_dist_to_2b, everything())

catcher_info
```

```{r sanity-checks}
# Should be zero — duplicate play keys would corrupt throw metrics
cat("Duplicate play keys:",
    steal_plays %>% count(game_string, play_per_game) %>%
      filter(n > 1) %>% nrow(), "\n")

cat("Steal plays:", nrow(steal_plays), "\n")
cat("Plays with a catcher throw:", sum(!is.na(catcher_info$throw_time)), "\n")
cat("Throws that arrived near 2B:", sum(catcher_info$throw_to_2b, na.rm = TRUE), "\n")

# Expected MiLB ranges: pop time ~1.9-2.3s, exchange ~0.7-1.0s, velo ~70-85 mph.
# Values far outside these suggest event-matching issues or noisy frames.
catcher_info %>%
  filter(throw_to_2b) %>%
  select(pop_time, exchange_time, max_throw_speed_mph, avg_throw_speed_mph) %>%
  summary()
```

## Pop time / exchange time / throw speed by outcome (Safe vs. Out)

```{r outcome-summary-safe}
# pop_time <= 3.5s screens out plays where the throw was deflected/overthrown
# and "receive_time" latched onto a much later ball-acquired event instead of
# a clean catch-and-tag (visible in ball_eventcode as bounce/deflection codes
# between the throw and the matched receive). Real pop times here top out
# around 2.05s, then jump straight to 4.15s+ for these mismatched plays.
cat("Safe\n")
catcher_info %>%
  filter(throw_to_2b, outcome == "Safe", pop_time <= 3.5) %>%
  select(pop_time, exchange_time, max_throw_speed_mph, avg_throw_speed_mph) %>%
  summary()
```

```{r outcome-summary-out}
# Includes "Out (inning-ending, verify)": manual review only relabeled one
# such play to a bare "Out", so restricting to outcome == "Out" alone would
# leave a single-row sample (all summary stats collapse to one value).
cat("Out\n")
catcher_info %>%
  filter(throw_to_2b, outcome %in% c("Out", "Out (inning-ending, verify)"),
         pop_time <= 3.5) %>%
  select(pop_time, exchange_time, max_throw_speed_mph, avg_throw_speed_mph) %>%
  summary()
```
